---
title: "Pigs example"
author: "Marc Comas-Cufí, Josep Antoni Martín-Fernandez and Glòria Mateu-Figueras"
date: "31 Dec 2015"
output: 
  html_document: 
    toc: yes
---


```{r, message=FALSE}
library(mixtools)
library(dplyr)
library(tidyr)
library(ggplot2)
library(mixpack)
```

# Data preprocessing

Pigs data has recorder the number of times different pigs were in bed, passaging or feedeing.

```{r}
if(!file.exists('Pigs.rda')){
  download.file('https://github.com/cran/zCompositions/blob/master/data/Pigs.rda?raw=true',
                destfile='Pigs.rda')
}
load('Pigs.rda')
Pigs %>% tbl_df
```

To fit a mixture of multinomial distributions and not to have problems of identifiability, we are going to restrict our analysis categories `BED`, `PASSAGE` and `FEEDER`.

```{r}
(pigs <- Pigs %>% dplyr::select(BED, PASSAGE, FEEDER)) %>% tbl_df
```

# Analysis

## Fitting the mixture of multinomial distributions

To fit the mixture of multinomial distributions we are going to use function `multmixEM()` from package `mixtools`.

```{r, results='hide'}
bic = function(loglik, k, n, dim) -2 * loglik + (k * dim - 1) * log(n)
set.seed(2)
suppressMessages(
  FITs <- lapply(2:15, function(K){
    FIT <- lapply(1:200, function(it, K){
      multmixEM(as.matrix(pigs), k = K)
      }, K)
    FIT[[which.min(sapply(FIT, function(fit) bic(fit$loglik, K, NROW(pigs), NCOL(pigs))))]]
  })
)
```

```{r, fig.width=5, fig.height=4}
df = lapply(FITs, function(fit){
  data.frame(
    K = length(fit$lambda),
    BIC = bic(fit$loglik, length(fit$lambda), NROW(pigs), NCOL(pigs))
  )
}) %>% bind_rows
plot(df, xlab='Number of components', ylab='BIC value', main='BIC value for different number of components')
```

We can see how the minimum BIC is obtained using a mixture of 9 multinomial distributions.

```{r}
df
```

The parameters of this mixtures are:

```{r}
fit = FITs[[5]]
fit$lambda
fit$theta
```

The groups can be represented by

```{r, fig.width=7, fig.height=4.5}
d = (pigs.g <- pigs %>% mutate(
  TOTAL = BED+PASSAGE+FEEDER,
  Comp = as.character(apply(fit$posterior, 1, which.max)))) %>%
  group_by(Comp) %>%
  summarise(
    BED = mean(BED/TOTAL),
    PASSAGE = mean(PASSAGE/TOTAL),
    FEEDER = mean(FEEDER/TOTAL)) %>%
  gather(key=Variable, value=Proportion, -Comp)

ggplot() +
  geom_bar(data = d, aes(x=Variable, y=Proportion, fill=Comp), stat = 'identity') +
  facet_wrap(~Comp, ncol=3) +
  theme(legend.position="none") + xlab("")
```

## Merging components

We continue our analyisis by building a sequential hierarchical partition. We consider $\omega = \tau_{i I_a}$ and $\lambda = \log^2 \left(\frac{ \tau_{iI_b} }{ \tau_{iI_a} }\right)$. The obtained sequential hiearchical partition is given by

```{r}
hp = get_hierarchical_partition(fit$posterior, omega = 'prop', lambda = 'coda.norm')
seq_partition = lapply(rev(hp), sapply, paste, collapse=',')
cat(paste(sapply(seq_partition, function(partition) paste(sprintf('{%s}', partition), collapse=',')), collapse='\n'))
```

In next figure, we can see the $S$-value's obtained in each step

```{r, fig.width=5, fig.height=4}
plot(attr(hp, 'S.value'), xlab='Clusters', ylab='S-value', main = 'S-values during the merging process')
```

Looking at $S$-value figure we can see that the drop in $S$-value happens between when considering 2 clusters instead of 3. This fact suggests that when merging from 3 to 2 clusters we are losing valuable information, and therefore, it seems reasonable to stop at 3 clusters.

  * 5 clusters

In the next figure we can see that components 5 and 6 are the first candidates to be merged. They both share the property that the pigs modeled by those two component have higher number in `BED` component and lower number in `FEEDER`.

```{r, fig.width=7, fig.height=4.5}
K = 5
clusters = mixpack::cluster_partition(fit$posterior, partition = hp[[K]] %>% setNames(., sapply(., function(lbl) paste(lbl, collapse='-'))))
d = pigs.g %>%
  mutate(
    Cluster = clusters
  ) %>% group_by(Cluster, Comp) %>%
  summarise(
    BED = mean(BED/TOTAL),
    PASSAGE = mean(PASSAGE/TOTAL),
    FEEDER = mean(FEEDER/TOTAL)) %>%
  gather(key=Variable, value=Proportion, -Cluster, -Comp)

ggplot() +
  geom_bar(data = d, aes(x=Variable, y=Proportion, fill=Comp), stat = 'identity', position=position_dodge()) +
  facet_wrap(~Cluster, ncol=3) +
  theme(legend.position="none") + xlab("")
```

  * 4 clusters

In the next figure we can see that components 3 and 4 are the second candidates to be merged. They both share the property that the pigs modeled by those two component have higher number in `FEEDER` component and lower number in `BED`.

```{r, fig.width=7, fig.height=4.5}
K = 4
clusters = mixpack::cluster_partition(fit$posterior, partition = hp[[K]] %>% setNames(., sapply(., function(lbl) paste(lbl, collapse='-'))))
d = pigs.g %>%
  mutate(
    Cluster = clusters
  ) %>% group_by(Cluster, Comp) %>%
  summarise(
    BED = mean(BED/TOTAL),
    PASSAGE = mean(PASSAGE/TOTAL),
    FEEDER = mean(FEEDER/TOTAL)) %>%
  gather(key=Variable, value=Proportion, -Cluster, -Comp)

ggplot() +
  geom_bar(data = d, aes(x=Variable, y=Proportion, fill=Comp), stat = 'identity', position=position_dodge()) +
  facet_wrap(~Cluster, ncol=3) +
  theme(legend.position="none") + xlab("")
```

  * 3 clusters
  
As suggested by $S$-values, it seems reasonable to stop at 3 clusters. In next figure we can see three classes of pigs. The first with higher proportion in `FEEDER`, the second with higher proportion in `BED` and the third with higher amount in `PASSAGE`.

```{r, fig.width=7, fig.height=2}
K = 3
clusters = mixpack::cluster_partition(fit$posterior, partition = hp[[K]] %>% setNames(., sapply(., function(lbl) paste(lbl, collapse='-'))))
d = pigs.g %>%
  mutate(
    Cluster = clusters
  ) %>% group_by(Cluster, Comp) %>%
  summarise(
    BED = mean(BED/TOTAL),
    PASSAGE = mean(PASSAGE/TOTAL),
    FEEDER = mean(FEEDER/TOTAL)) %>%
  gather(key=Variable, value=Proportion, -Cluster, -Comp)

ggplot() +
  geom_bar(data = d, aes(x=Variable, y=Proportion, fill=Comp), stat = 'identity', position=position_dodge()) +
  facet_wrap(~Cluster, ncol=3) +
  theme(legend.position="none") + xlab("")
```
